---
title: "Apparent cumulative deficit at FLUXNET sites"
author: "Beni Stocker"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyr)
source("./R/get_obs_bysite_fluxnet2015.R")
```

## Idea

This derives the apparent cumulative water deficit, based on observed ET at FLUXNET sites. The extreme values with a return period of N years is derived as a measure for the magnitude of droughts to which vegetation is expected to be adapted to. 

## Get ET data

In order to do a gapfilling using variables that are known to determine fluxes, we select predictors that are commontly used for estimating potential evapotranspiration using the Penman-Monteith Equation. These are:

- $R_n$ net radiation
- $G$ ground heat flux (ignored)
- $S$ sum of all storage fluxes (ignored)
- $\rho$ air density (`air.density()`)

  - air temperature
  - atmospheric pressure

- $D_a$ VPD
- $G_\text{ah}$ aerodynamic conductance to heat transfer

  - $G_\text{am}$ aerodynamic conductance (`aerodynamic.conductance()`)
  
    - air temperature
    - pressure
    - wind speed at measurement height $u(z_r)$
    - Friction velocity $u_\ast$
    - sensible heat flux $H$
    - local parameters (not needed for the simplest form using `Rb_model = "Thom_1972"`)
    
      - instrument height $z_r$
      - canopy height $z_h$
      - zero-plane displacement height $d$
      - roughness length for momentum $z_{0m}$
      - characteristic leaf dimension $D1$
      
  - $R_\text{bh}$ canopy boundary layer resistance
  
    - Friction velocity $u_\ast$

- $s$ slope of the saturation vapor pressure curve (`Esat.slope()`)

  - air temperature

- $\gamma$ psychrometric constant (`psychrometric.constant()`)

  - air temperature
  - atmospheric pressure
  
- $G_\text{sw}$ Canopy conductance. Unknown, expected to respond to the following variables

  - soil moisture
  - PPFD
  - VPD
  
More detailed formulations of the atmospheric conductance formulations account for LAI, leaf width. 

### Daily data

We use filtered daily data and conduct a filtering, similar as applied by the function `filter.data()` from the bigleaf package. 

- Exclude rainy days for ET estimate
- Low $u_\ast$

In summary, we selected the following predictors to gapfill ET:

- air temperature, day and night
- atmospheric pressure
- net radiation
- sensible heat flux
- wind speed
- PPFD
- VPD
- friction velocity

Read data from file
```{r warning=FALSE}
getvars <- c( 
  "TA_F_NIGHT", 
#  "TA_F_NIGHT_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "TA_F_DAY", # degC
#  "TA_F_DAY_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "SW_IN_F", # W m-2
#  "SW_IN_F_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "VPD_F", # hPa
#  "VPD_F_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "PA_F", # kPa
#  "PA_F_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "P_F",  # mm
#  "P_F_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "WS_F",  # m s-1
#  "WS_F_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "USTAR", # m s-1
#  "USTAR_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "NETRAD", # W m-2
#  "NETRAD_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "H_F_MDS", # W m-2 -> J m-2 d-1 or H_CORR (corrected by energy balance closure factor)?
#  "H_F_MDS_QC",  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  "LE_F_MDS"  # W m-2 -> J m-2 d-1
#  "LE_F_MDS_QC"  # DD: % measured or good qual gapfilled; HH: 0=measured, 1=good qual gapfill, 2=medium, 3=poor
  )

ddf <- get_obs_bysite_fluxnet2015( 
  sitename="AU-How", 
  path_fluxnet2015="/alphadata01/bstocker/data/FLUXNET-2015_Tier1/20160128/point-scale_none_1d/original/unpacked/", 
  timescale="d", 
  getvars=getvars,
  threshold_GPP=0, threshold_LE=0.5, threshold_H=0.5, threshold_SWC=1.0, threshold_WS=0.3, threshold_USTAR=0.5, threshold_T=0.5
  ) 
```

Filter data using the `filter.data()` function from the bigleaf package.
```{r eval=FALSE}
ddf_clean <- bigleaf::filter.data(
  data = ddf,
  quality.control = TRUE,
  vars.qc = c("temp_NIGHT", "temp_DAY", "swin", "vpd", "patm", "prec", "wspeed", "ustar", "netrad", "sensibleh", "latenth"),
  quality.ext = "_QC",
  good.quality = c(50,100),
  missing.qc.as.bad = TRUE )
```

Plot some of the data that has just been loaded.
```{r}
ddf %>% 
  ggplot(aes(date, latenth)) +
  geom_line()

ddf %>% 
  ggplot(aes(date, sensibleh)) +
  geom_line()

ddf %>% 
  ggplot(aes(date, wspeed)) +
  geom_line()

ddf %>% 
  ggplot(aes(date, ustar)) +
  geom_line()

ddf %>% 
  ggplot(aes(date, SWC_F_MDS_1)) +
  geom_line()
```


## Gapfill data

```{r}
source("R/gapfill_nn.R")
predictors <- names(ddf)
predictors <- predictors[!(grepl("_QC", predictors))]
predictors <- predictors[-which(predictors=="latenth" || predictors=="date")]

ddf_gapfilled <- gapfill_nn(ddf, predictors = predictors, nam_target = "latenth", package = "caret")
```

Look at what was done.
```{r}
ggplot() +
  geom_line(data=ddf_gapfilled, aes(date, latenth), color="blue") +
  geom_line(data=ddf, aes(date, latenth), color="black")
```





















