---
title: "ET decay model"
author: "Beni Stocker"
date: "4/26/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Linear model

Assume a 60-days dry down event with the following setup:

- The initial water available to plants across the rooting zone, $S_0$ is 100 mm.
- ET (termed $T$, since it's only transpiration here) is a linear function of the remaining water stored $S_t$ and is independent of VPD.
$$
T = \alpha S/S_0
$$
- The change in plant-available water storage is $\Delta S=T$.

This leads to an exponential decay of both $S$ and $D$ with time. We can set $\alpha = 0.01$ d $^{-1}$.
```{r}
s0 <- 100
alpha <- 1
ntsteps <- 500
s_t <- rep(NA, ntsteps)
t_t <- rep(NA, ntsteps)
s_t[1] <- s0 

for (idx in 1:(ntsteps-1)){
  t_t[idx] <- alpha * s_t[idx] / s0
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}

df_lin <- tibble( time = 1:ntsteps, soilm = s_t, transp = t_t ) 

df_lin %>% 
  ggplot(aes(time, soilm)) + 
  geom_line()

df_lin %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, transp)) + 
  geom_line()

df_lin %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(100-soilm, transp)) + 
  labs(x=expression(integral(ET))) +
  geom_line()
```

### Effect of rooting zone water storage capacity

Let's compare the same relationship as above but with two different $S_0$.
```{r}
s0 <- 50
s_t[1] <- s0 
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- alpha * s_t[idx] / s0
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}


df_lin_deep <- df_lin %>% 
  mutate(rzwsc = "deep", cwd = cumsum(transp))

df_lin_shallow <- tibble(time = 1:ntsteps, soilm= s_t, transp = t_t) %>% 
  mutate(rzwsc = "shallow", cwd = cumsum(transp))

df_lin <- bind_rows(df_lin_deep, df_lin_shallow)

df_lin %>% 
  ggplot(aes(x = time, y = soilm, color = rzwsc)) + 
  geom_line()

df_lin %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, transp, color = rzwsc)) + 
  geom_line()

df_lin %>%
  tidyr::drop_na() %>% 
  ggplot(aes(x = cwd, y = transp, color = rzwsc)) +
  labs(x=expression(integral(ET))) +
  geom_line()
```


The question is: How does VPD affect the ET decline?

## VPD-enabled model, linear

We can formulate a model for $T$ as a function of VPD (termed $D$) and canopy conductance $G_s$.
$$
T = G_s D
$$
$G_s$ itself is a function of the remaining soil water content $S$. We can assume a linear model with $G_s = 0$ for $S = 0$:
$$
G_s = \beta S
$$
To get comparable numbers to the example above, let's assume that under under well-watered conditions ($S = 100$ mm) and a VPD of 1000 Pa, we should have a transpiration of 1 mm d$^{-1}$. Hence, we $\beta = 10^{-5}$.
$$
T = \beta S D
$$

How does $T$ evolve under constant $D$?
```{r}
s0 <- 100
beta <- 1e-5
ntsteps <- 500
d_t <- rep(1000, ntsteps)
s_t <- rep(NA, ntsteps)
t_t <- rep(NA, ntsteps)
s_t[1] <- s0
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- beta * s_t[idx] * d_t[idx]
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}
df_vpd <- tibble( time = 1:ntsteps, soilm_vpd_const = s_t, transp_vpd_const = t_t ) 
df_vpd %>% 
  ggplot(aes(time, soilm_vpd_const)) + 
  geom_line()

df_vpd %>% 
  tidyr::drop_na() %>% 
  ggplot(aes(time, transp_vpd_const)) + 
  geom_line()
```

Ok. That's the same. As expected because transpiration is effectively just a function of the $S$, equivalent to the linear model above. 

But what happens when $D$ changes over time? Let's assume a linear increase from 1000 Pa to 4000 Pa.
```{r}
s0 <- 100
beta <- 1e-5
ntsteps <- 500
d_t <- rep(NA, ntsteps)
d_t[1] <- 1000
d_t[ntsteps] <- 4000
d_t <- approx(1:ntsteps, d_t, xout=1:ntsteps)$y
s_t <- rep(NA, ntsteps)
t_t <- rep(NA, ntsteps)
s_t[1] <- s0
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- beta * s_t[idx] * d_t[idx]
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}
df_vpd <- tibble( time = 1:ntsteps, soilm_vpd_lin = s_t, transp_vpd_lin = t_t ) %>% left_join(df_vpd, by="time")
df_vpd %>% 
  ggplot(aes(time)) + 
  geom_line(aes(y=soilm_vpd_const, color="soilm_vpd_const"), linetype="dashed") +
  geom_line(aes(y=soilm_vpd_lin, color="soilm_vpd_lin"))

df_vpd %>% 
  ggplot(aes(time)) + 
  geom_line(aes(y=transp_vpd_const, color="transp_vpd_const"), linetype="dashed") +
  geom_line(aes(y=transp_vpd_lin, color="transp_vpd_lin"))
```

Hmmm. That's not the same.

## VPD-enabled model, sigmoidal

### Constant VPD

Instead of a linear response of $G_s$ to $S$, we can assume a sigmoidal response that looks (somewhat more realistically) like this:
```{r}
sigm <- function(x){0.001/(1+exp(-(x/5-5)))}
df_sigm <- tibble(soilm=0:100) %>% 
  rowwise() %>% 
  mutate(gs = sigm(soilm))
df_sigm %>% 
  ggplot(aes(soilm, gs)) +
  geom_line()
```

Now, let's look at the evolution of soil moisture and transpiration during dry-down events. First, with constant VPD.
```{r}
beta <- 1e-5
ntsteps <- 500
d_t <- rep(1000, ntsteps)
s_t <- rep(NA, ntsteps)
t_t <- rep(NA, ntsteps)
s_t[1] <- s0
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- sigm(s_t[idx]) * d_t[idx]
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}
df_vpd <- tibble( time = 1:ntsteps, soilm_vpd_const_sigm = s_t, transp_vpd_const_sigm = t_t ) %>% left_join(df_vpd, by="time")
df_vpd %>% 
  ggplot(aes(time)) + 
  geom_line(aes(y=soilm_vpd_const, color="soilm_vpd_const"), linetype="dashed") +
  geom_line(aes(y=soilm_vpd_const_sigm, color="soilm_vpd_const_sigm"))

df_vpd %>% 
  ggplot(aes(time)) + 
  geom_line(aes(y=transp_vpd_const, color="transp_vpd_const"), linetype="dashed") +
  geom_line(aes(y=transp_vpd_const_sigm, color="transp_vpd_const_sigm"))

df_vpd %>% 
  ggplot() + 
  geom_line(aes(x=100-soilm_vpd_const, y=transp_vpd_const, color="transp_vpd_const"), linetype="dashed") +
  geom_line(aes(x=100-soilm_vpd_const_sigm, y=transp_vpd_const_sigm, color="transp_vpd_const_sigm"))
```

### Increasing VPD

```{r}
beta <- 1e-5
ntsteps <- 500
d_t <- rep(NA, ntsteps)
d_t[1] <- 1000
d_t[ntsteps] <- 4000
d_t <- approx(1:ntsteps, d_t, xout=1:ntsteps)$y
s_t <- rep(NA, ntsteps)
t_t <- rep(NA, ntsteps)
s_t[1] <- s0
for (idx in 1:(ntsteps-1)){
  t_t[idx] <- sigm(s_t[idx]) * d_t[idx]
  s_t[idx + 1] <- s_t[idx] - t_t[idx]
}
df_vpd <- tibble( time = 1:ntsteps, soilm_vpd_lin_sigm = s_t, transp_vpd_lin_sigm = t_t ) %>% left_join(df_vpd, by="time")
df_vpd %>% 
  ggplot(aes(time)) + 
  geom_line(aes(y=soilm_vpd_const, color="soilm_vpd_const"), linetype="dashed") +
  geom_line(aes(y=soilm_vpd_lin_sigm, color="soilm_vpd_lin_sigm"))

df_vpd %>% 
  ggplot(aes(time)) + 
  geom_line(aes(y=transp_vpd_const, color="transp_vpd_const"), linetype="dashed") +
  geom_line(aes(y=transp_vpd_lin_sigm, color="transp_vpd_lin_sigm"))
```